Practical Optional Types for Clojure

### 2015-02-27 formal-model.tex
overview
isa? formal

The particular part of the environment we reference in a positive or negative
type proposition is called the \emph{object}.
Every expression has an object that is either \emph{empty}, written \emptyobject{}, 
which says 
this expression is not known to evaluate to a particular component
  of the current environment, or a 
variable with some \emph{path}, written \path{\pathelem{}}{\x{}},
that exactly indicates how the value of this
expression can be derived from the current environment.
Type propositions can only reference non-empty objects.

@@

    %{\TDo}
    %{\TClass}
    {\TIf}
    {\TLet}
    {\TConst}

    %{\TKw}

@@

Recall example~\ref{example:doexception}.
\begin{minted}{clojure}
      (do (if (number? x) nil (throw (Exception.)))
          (inc x))
\end{minted}

@@

Before tackling multimethods we model their core dispatch mechanism
\isaliteral{}, a subclassing test for classes, otherwise an equality test.

\begin{minted}{clojure}
(isa? Keyword Object) ;=> true
(isa? Object Keyword) ;=> false
(isa? :a :a) ;=> true
(isa? :a :b) ;=> false
(isa? :a Keyword) ;=> false
(isa? (class :a) Keyword) ;=> true
\end{minted}

B-IsA with the metafunction \isaopsemliteral{} reflect these semantics.

@@

For the typing rules, we add T-Kw and T-Class for type checking keyword and class literals
as singleton types, and T-IsA for \isaliteral{} expressions, which 
uses the \emph{object} of the left argument and the \emph{type} of the right argument
to infer the final proposition set.

$$
\begin{array}{ll}
\vcenter{\hbox{
    \TIsA{}}} &

  \vcenter{\hbox{$
    \begin{array}{l}
      {\TKw}\\\\
      {\TClass}
    \end{array}$}}
\end{array}
$$

The definition of \isacompareliteral{} (figure~\ref{main:figure:mmsyntax}) is deliberately conservative.
The first line considers the case where the object of the left argument
is a non-empty path ending in \classpe{} and the type of the right argument is a singleton class.
The simplest example

\begin{minted}{clojure}
  (isa? (class x) Keyword)
\end{minted}

has the proposition set \isacompare{\s{}}{\path{\classpe{}}{\x{}}}{\Value{\Keyword}}{\filterset{\isprop{\Keyword}{\x{}}}{\notprop{\Keyword}{\x{}}}},
that is if the expression is true then \clj{x} is a keyword, otherwise \clj{x} is not a keyword.

We can rewrite example~\ref{example:conditionalflow} as
\inputminted[firstline=5]{clojure}{code/demo/src/demo/isa_nil.clj}

because the then proposition is {\isprop{\Class}{\path{\classpe{}}{\x{}}}}, so
we know {\isprop{\Object}{\x{}}} by L-Update, and combined with the fact {\isprop{\Union{\Nil{}}{\Number}}{\x{}}}
we can infer {\isprop{\Number}{\x{}}} by L-Update.

The second \isacompareliteral{} line handles the equality mode of \isaliteral{}
with a non-class singleton type as the second argument.

@@

%Substitution is used to ensure well formed propositions.

The final line gives up---we cannot distinguish between equality and subclassing mode.

>>>>

The particular part of the environment we reference in a
type proposition is called the \emph{object}.
Every expression has an object that is either \emph{empty}, written \emptyobject{}, 
which says 
this expression is not known to evaluate to a particular component
  of the current environment, or a 
variable with some \emph{path}, written \path{\pathelem{}}{\x{}},
that exactly indicates how the value of this
expression can be derived from the current environment.
Type propositions can only reference non-empty objects.

@@

    %{\TDo}
    %{\TClass}
    {\TIf}
    {\TLet}
    \begin{array}{c}
      {\TConst}\\\\
      {\TKw}\\\\
      {\TClass}
    \end{array}

@@

Recall example~\ref{example:doexception}.
\begin{minted}{clojure}
...  (do (if (number? x) nil (throw (Exception.)))
         (inc x)) ...
\end{minted}

@@

We now consider the core dispatch mechanism for multimethods. 
Recalling the examples in section~\ref{sec:multioverview},
\isaliteral{} is
a subclassing test for classes, otherwise an equality test---we do not model
the semantics for vectors.

The main interest for the typing rule is \isacompareliteral{} 
(figure~\ref{main:figure:mmsyntax}).

\begin{mathpar}
  \TIsA{}
\end{mathpar}

To demonstrate,
\mintinline{clojure}{(isa? (class x) Keyword)}
has the proposition set \isacompare{\s{}}{\path{\classpe{}}{\x{}}}{\Value{\Keyword}}{\filterset{\isprop{\Keyword}{\x{}}}{\notprop{\Keyword}{\x{}}}}.

B-IsA models the semantics. \isaopsemliteral{} explicitly handles classes in the second case.

@@

%The definition of \isacompareliteral{} (figure~\ref{main:figure:mmsyntax}) is deliberately conservative.
%The first line considers the case where the object of the left argument
%is a non-empty path ending in \classpe{} and the type of the right argument is a singleton class.

@@

### 2015-02-27 overview.tex
isa? formal

\paragraph{Semantics} \clj{get} applied to a hash-map and a key returns the value
associated with the key in the map, or \clj{nil} if there is none.
Keywords are functions that apply \clj{get} to their argument
and itself.
\clj{inc} and \clj{dec} increment and decrement numbers;
\clj{assoc} and \clj{dissoc} associate and dissociate entries
from immutable maps.
The \clj{=} function returns true if both arguments are equal.
\clj{number?} is
a Java \java{instanceof} test of \clj{java.lang.Number}.
\clj{nil} and \clj{false} are the false values.

@@

\paragraph{Typed Clojure Macros}
The \clj{defalias} macro defines a type abbreviation which can reference itself recursively.
The typed \clj{fn} has optional annotations by following a parameter
or binding vector by \clj{:-} and a type.

@@

\citet{TF10}
presented Typed Racket with occurrence typing,
a technique for deriving type information from conditional control flow.
They introduce the concept of occurrence typing 
with the following example.

@@

We chose to build on the ideas and implementation
of Typed Racket to implement a type system targeting Clojure for several reasons.
Initially, the similarities between Racket and Clojure drew us to
investigate the effectiveness of repurposing occurrence typing
for a Clojure type system---both languages share a Lisp heritage,
similar standard functions 
(for instance \clj{map}
in both languages is variable-arity)
and idioms.
While Typed Racket is gradually typed and has sophisticated
dynamic semantics for cross-language interaction, we 
chose to first implement
the static semantics
with the hope to extend Typed Clojure to be gradually typed at a future date.
Finally,
Typed Racket's combination of bidirectional checking
and occurrence typing presents a successful model for 
type checking dynamically typed programs without compromising
soundness, where other approaches like success typing~\cite{todo-dialyzer-paper}
or soft typing~\cite{successtyping-todo}
fail to prove strong properties about programs
and have proved too complicated in practice respectively.

@@

Along with conditional control flow,
Clojure programmers rely on \emph{exceptions}
to assert type-related invariants, such as preventing
null-pointer exceptions.

@@

To check this example,
occurrence typing 
automatically
assumes
\clj{x} is a number when checking the second \clj{do} subexpression
based on the first subexpression.
\footnote{We omit \clj{ns} forms for the rest of the examples, see <this> github repo for full working examples}
We model this formally (section~\ref{sec:doformal}) and prove
null-pointer exceptions are impossible in typed code (section~\ref{sec:metatheory}).

@@

The next example is extracted from a Typed Clojure production system at CircleCI
(section~\ref{sec:casestudy} presents a case study).

@@

@@

A multimethod in Clojure is a function that contains a dispatch
function and methods. Multimethods are created with {\clj{defmulti}}.
\inputminted[firstline=5,lastline=7]{clojure}{code/demo/src/demo/rep.clj}
The multimethod \clj{path} has type \clj{[Any -> String]}, an initially empty \emph{dispatch table}
and \emph{dispatch function} \clj{class}, a function that returns the argument's class or \clj{nil} if none.
There are currently no methods so \clj{path} will fail at runtime.
\begin{minted}{clojure}
  (path (File. "dir/a"));=> IllegalArgumentException
\end{minted}

>>>>

\paragraph{Semantics} \clj{get} applied to a hash-map and a key returns the value
associated with the key in the map, or \clj{nil} if there is none.
Keywords are functions that apply \clj{get} to their argument
and itself.
\clj{inc} and \clj{dec} increment and decrement numbers;
\clj{assoc} and \clj{dissoc} associate and dissociate entries
from immutable maps.
\clj{class} returns the argument's class, or \clj{nil} if passed \clj{nil}.
The \clj{=} function returns true if both arguments are equal.
\clj{number?} is
a Java \java{instanceof} test of \clj{java.lang.Number}.
\clj{nil} and \clj{false} are the false values.

@@

\paragraph{Typed Clojure Macros}
The \clj{defalias} macro defines a type abbreviation which can reference itself recursively.
The typed \clj{fn} supports optional annotations by 
adding
\clj{:-} and a type after a parameter
position
or binding vector 
to annotate parameter types
and return types respectively.

@@

\citet{TF10}
presented Typed Racket with occurrence typing,
a technique for deriving type information from conditional control flow.
They introduced the concept of occurrence typing 
with the following example.

@@

We chose to build on the ideas and implementation
of Typed Racket to implement a type system targeting Clojure for several reasons.
Initially, the similarities between Racket and Clojure drew us to
investigate the effectiveness of repurposing occurrence typing
for a Clojure type system---both languages share a Lisp heritage,
similar standard functions 
(for instance \clj{map}
in both languages is variable-arity)
and idioms.
While Typed Racket is gradually typed and has sophisticated
dynamic semantics for cross-language interaction, we 
chose to first implement
the static semantics
with the hope to extend Typed Clojure to be gradually typed at a future date.
Finally,
Typed Racket's combination of bidirectional checking
and occurrence typing presents a successful model for 
type checking dynamically typed programs without compromising
soundness, which is appealing over success typing~\cite{todo-dialyzer-paper}
which cannot prove strong properties about programs
and soft typing~\cite{successtyping-todo}
which has proved too complicated in practice.

@@

Along with conditional control flow,
Clojure programmers rely on \emph{exceptions}
to assert type-related invariants.

@@

In this example a \clj{throw} expression guards \clj{(inc x)}
from being evaluated if \clj{x} is \clj{nil}, preventing
a possible null-pointer exception.

To check this example,
occurrence typing 
automatically
assumes
\clj{x} is a number when checking the second \clj{do} subexpression
based on the first subexpression.
\footnote{See \url{https://github.com/typedclojure/examples}
  for full examples. From here we omit \clj{ns} forms.}
We model this formally (section~\ref{sec:doformal}) and prove
null-pointer exceptions are impossible in typed code (section~\ref{sec:metatheory}).

@@

The next example is extracted from a production system at CircleCI
(section~\ref{sec:casestudy} presents a case study).

@@

The next example is the same except we use the \clj{:absent-keys}
HMap option
and
utilize destructuring.

@@

\label{sec:multioverview}

A multimethod in Clojure is a function with a \emph{dispatch
function} and a \emph{dispath table} of methods. Multimethods are created with {\clj{defmulti}}.
\inputminted[firstline=5,lastline=7]{clojure}{code/demo/src/demo/rep.clj}
The multimethod \clj{path} has type \clj{[Any -> (U nil String)]}, an initially empty \emph{dispatch table}
and \emph{dispatch function} \clj{class}.
There are currently no methods so \clj{path} will fail at runtime.
\begin{minted}{clojure}
  (path (File. "dir/a"));=> IllegalArgumentException
\end{minted}
